---
title: "Modeling nonlinear relationships"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(scales)
library(patchwork)
library(lubridate)

source("R/render/inline_format.R")
source("R/pkg/model_table.R")

theme_set(theme_minimal(base_size = 13))

# Austin: logged response with area and ZIP code
austin <- read_csv(
  "data/austin_houses/austin_houses_subset.csv",
  show_col_types = FALSE
) |>
  mutate(zipcode = relevel(factor(zipcode), ref = "78731"))

fit_austin_raw <- lm(
  latest_price ~ living_area_sq_ft + zipcode,
  data = austin
)
fit_austin_log <- lm(
  log(latest_price) ~ living_area_sq_ft + zipcode,
  data = austin
)
fit_austin_rich <- lm(
  log(latest_price) ~ living_area_sq_ft + zipcode + age +
    num_of_bathrooms + lot_size_sq_ft + has_view + num_of_bedrooms +
    garage_spaces + num_of_stories,
  data = austin
)

austin_raw_diag <- austin |>
  mutate(
    fitted = predict(fit_austin_raw),
    residual = resid(fit_austin_raw)
  )
austin_log_diag <- austin |>
  mutate(
    fitted = predict(fit_austin_log),
    residual = resid(fit_austin_log)
  )

austin_log_coef <- coef(fit_austin_log)
austin_log_rse <- sigma(fit_austin_log)
austin_base_adj_r2 <- summary(fit_austin_log)$adj.r.squared
austin_rich_adj_r2 <- summary(fit_austin_rich)$adj.r.squared
area_100_pct <- 100 * (exp(100 * austin_log_coef[["living_area_sq_ft"]]) - 1)
zip_78721_pct <- 100 * (exp(austin_log_coef[["zipcode78721"]]) - 1)
zip_78723_pct <- 100 * (exp(austin_log_coef[["zipcode78723"]]) - 1)
austin_resid_ratio <- exp(austin_log_rse)

austin_prediction_data <- tibble(
  living_area_sq_ft = 2000,
  zipcode = factor(c("78731", "78721", "78723"), levels = levels(austin$zipcode))
) |>
  mutate(
    fitted_log_price = predict(fit_austin_log, newdata = pick(everything())),
    predicted_price = exp(fitted_log_price)
  )

austin_area_examples <- tibble(
  living_area_sq_ft = c(2000, 2100),
  zipcode = factor("78731", levels = levels(austin$zipcode))
) |>
  mutate(
    fitted_log_price = predict(fit_austin_log, newdata = pick(everything())),
    predicted_price = exp(fitted_log_price)
  )

area_log_difference <- diff(austin_area_examples$fitted_log_price)[[1]]
area_fitted_ratio <- exp(area_log_difference)

austin_zip_examples <- tibble(
  living_area_sq_ft = 2000,
  zipcode = factor(c("78731", "78723"), levels = levels(austin$zipcode))
) |>
  mutate(
    fitted_log_price = predict(fit_austin_log, newdata = pick(everything())),
    predicted_price = exp(fitted_log_price)
  )

zip_log_difference <- diff(austin_zip_examples$fitted_log_price)[[1]]
zip_fitted_ratio <- exp(zip_log_difference)

stopifnot(
  area_log_difference > 0,
  area_fitted_ratio > 1,
  zip_log_difference < 0,
  zip_fitted_ratio < 1,
  zip_78721_pct < 0,
  zip_78723_pct < 0
)

# Kroger cheese demand
kroger <- read_csv("data/kroger_cheese/kroger.csv", show_col_types = FALSE)
fit_kroger_raw <- lm(vol ~ price, data = kroger)
fit_kroger_log <- lm(log(vol) ~ log(price), data = kroger)

kroger_raw_diag <- kroger |>
  mutate(
    fitted = predict(fit_kroger_raw),
    residual = resid(fit_kroger_raw)
  )
kroger_log_diag <- kroger |>
  mutate(
    fitted = predict(fit_kroger_log),
    residual = resid(fit_kroger_log)
  )

kroger_coef <- coef(fit_kroger_log)
kroger_r2 <- summary(fit_kroger_log)$r.squared
kroger_price_ratio <- 2.50 / 2.00
kroger_volume_ratio <- kroger_price_ratio^kroger_coef[["log(price)"]]

kroger_curve <- tibble(
  price = seq(min(kroger$price), max(kroger$price), length.out = 300)
) |>
  mutate(
    raw_fit = predict(fit_kroger_raw, newdata = pick(everything())),
    log_log_fit = exp(predict(fit_kroger_log, newdata = pick(everything())))
  )

model_colors <- setNames(
  RColorBrewer::brewer.pal(3, "Set2")[1:2],
  c("Raw linear", "Log-log")
)

# ERCOT daily data and specified polynomial models
load_ercot_daily <- function(years) {
  loads <- bind_rows(
    read_csv("data/ercot/ercotproject/Data/load_1.csv", show_col_types = FALSE),
    read_csv("data/ercot/ercotproject/Data/load_2.csv", show_col_types = FALSE),
    read_csv("data/ercot/ercotproject/Data/load_3.csv", show_col_types = FALSE)
  ) |>
    filter(LOCATION == "North Central (DFW)") |>
    mutate(datetime = parse_date_time(DATEHOUR, orders = c("ymd HM", "ymd HMS"))) |>
    filter(year(datetime) %in% years)

  weather <- read_csv(
    "data/ercot/ercotproject/Data/weather_ncent.csv",
    show_col_types = FALSE
  ) |>
    mutate(datetime = parse_date_time(DATEHOUR, orders = c("ymd HM", "ymd HMS"))) |>
    filter(year(datetime) %in% years)

  loads |>
    inner_join(
      weather,
      by = c("datetime", "LOCATION"),
      relationship = "many-to-many"
    ) |>
    filter(!is.na(HourlyDryBulbTemperature), !is.na(LOAD)) |>
    distinct(datetime, .keep_all = TRUE) |>
    mutate(date = as.Date(datetime)) |>
    summarize(
      temp = mean(HourlyDryBulbTemperature),
      load = mean(LOAD),
      .by = date
    ) |>
    arrange(date)
}

degrees <- c(1, 2, 3, 5, 9)
ercot_2018 <- load_ercot_daily(2018)

fit_degrees <- function(data, y, x, degrees_to_fit = degrees) {
  forms <- lapply(degrees_to_fit, function(k) {
    reformulate(sprintf("poly(%s, %d)", x, k), response = y)
  })
  setNames(
    lapply(forms, function(f) lm(f, data = data)),
    paste0("degree ", degrees_to_fit)
  )
}

summarize_fits <- function(fits) {
  tibble(
    degree = degrees,
    r_squared = map_dbl(fits, ~ summary(.x)$r.squared),
    adj_r_squared = map_dbl(fits, ~ summary(.x)$adj.r.squared),
    residual_sd = map_dbl(fits, sigma)
  )
}

fitted_curves <- function(fits, data, x, n_grid = 400) {
  grid <- seq(min(data[[x]]), max(data[[x]]), length.out = n_grid)
  map2_dfr(fits, names(fits), function(fit, name) {
    newdata <- setNames(tibble(grid), x)
    tibble(
      x = grid,
      y = predict(fit, newdata = newdata),
      degree = factor(sub("degree ", "", name), levels = as.character(degrees))
    )
  })
}

ercot_fits <- fit_degrees(ercot_2018, "load", "temp")
ercot_fit_stats <- summarize_fits(ercot_fits)
ercot_curves <- fitted_curves(ercot_fits, ercot_2018, "temp")
ercot_degree1_diag <- ercot_2018 |>
  mutate(
    fitted = predict(ercot_fits[["degree 1"]]),
    residual = resid(ercot_fits[["degree 1"]])
  )
ercot_degree2_diag <- ercot_2018 |>
  mutate(
    fitted = predict(ercot_fits[["degree 2"]]),
    residual = resid(ercot_fits[["degree 2"]])
  )
ercot_degree3_diag <- ercot_2018 |>
  mutate(
    fitted = predict(ercot_fits[["degree 3"]]),
    residual = resid(ercot_fits[["degree 3"]])
  )
ercot_degree5_diag <- ercot_2018 |>
  mutate(
    fitted = predict(ercot_fits[["degree 5"]]),
    residual = resid(ercot_fits[["degree 5"]])
  )

ercot_temp_range <- range(ercot_2018$temp)
ercot_extrap_grid <- tibble(
  temp = seq(ercot_temp_range[1] - 8, ercot_temp_range[2] + 8, length.out = 500)
)
ercot_extrap_curves <- map2_dfr(ercot_fits, names(ercot_fits), function(fit, name) {
  ercot_extrap_grid |>
    mutate(
      load = predict(fit, newdata = pick(everything())),
      degree = factor(sub("degree ", "", name), levels = as.character(degrees))
    )
})

# Austin 78744 polynomial example, prepared exactly as specified
austin_78744 <- read_csv(
  "data/austin_houses/austin_houses.csv",
  show_col_types = FALSE
) |>
  filter(zipcode == 78744, is.finite(age), is.finite(latest_price)) |>
  filter(latest_price < quantile(latest_price, 0.99)) |>
  rename(price = latest_price)

austin_poly_fits <- fit_degrees(austin_78744, "price", "age")
austin_poly_stats <- summarize_fits(austin_poly_fits)
austin_poly_curves <- fitted_curves(austin_poly_fits, austin_78744, "age")

austin_max_age <- max(austin_78744$age)
austin_extrap_grid <- tibble(
  age = seq(min(austin_78744$age), austin_max_age + 8, length.out = 500)
) |>
  mutate(
    degree_2 = predict(austin_poly_fits[["degree 2"]], newdata = pick(everything())),
    degree_9 = predict(austin_poly_fits[["degree 9"]], newdata = pick(everything()))
  ) |>
  pivot_longer(
    starts_with("degree_"),
    names_to = "degree",
    values_to = "price"
  ) |>
  mutate(
    degree = recode(degree, degree_2 = "Degree 2", degree_9 = "Degree 9"),
    degree = factor(degree, levels = c("Degree 2", "Degree 9"))
  )

austin_degree9_at_68 <- predict(
  austin_poly_fits[["degree 9"]],
  newdata = tibble(age = 68)
)[[1]]

stopifnot(
  austin_max_age < 68,
  austin_degree9_at_68 < 0
)

ercot_table <- ercot_fit_stats |>
  transmute(
    Degree = degree,
    `$R^2$` = number(r_squared, accuracy = 0.001),
    `Adjusted $R^2$` = number(adj_r_squared, accuracy = 0.001),
    `Residual SD (MW)` = comma(residual_sd, accuracy = 1)
  )

austin_poly_table <- austin_poly_stats |>
  transmute(
    Degree = degree,
    `$R^2$` = number(r_squared, accuracy = 0.001),
    `Adjusted $R^2$` = number(adj_r_squared, accuracy = 0.001),
    `Residual SD` = dollar(residual_sd, accuracy = 1)
  )
```

Linear regression models can fit certain *nonlinear* relationships too. Interaction terms already made fitted relationships nonlinear in the predictors by including their products. Transforming covariates, responses, or both lets us represent other nonlinear relationships.

## Log transformations in regression

When we checked diagnostics for the Austin housing data, the spread in our prediction errors grew with the fitted value. That pattern suggests trying a logarithmic scale for price. As we saw in @sec-logarithms, equal vertical distances on a log scale represent equal ratios rather than equal dollar differences.

```{r}
#| label: fig-austin-scales
#| fig-cap: "Austin sale price against living area. Left: price in dollars. Right: log price compresses the larger dollar gaps among higher-priced homes."
#| fig-width: 10
#| fig-height: 4.5

p_austin_raw <- ggplot(austin, aes(living_area_sq_ft, latest_price)) +
  geom_point(color = "gray50", alpha = 0.35, size = 0.9) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "K")) +
  labs(x = "Living area (square feet)", y = "Sale price", title = "Price in dollars")

p_austin_log <- ggplot(austin, aes(living_area_sq_ft, latest_price)) +
  geom_point(color = "gray50", alpha = 0.35, size = 0.9) +
  scale_x_continuous(labels = comma) +
  scale_y_log10() +
  labs(x = "Living area (square feet)", y = "Sale price (log scale)", title = "Log price") 

p_austin_raw + p_austin_log
```

Even without a fitted line, we can already see that the spread *around* the line of best fit will be more uniform on the right. 


<!-- The raw-price model is

TODO: I DON"T WANT EXTRA GREEK COEFS. ALWAYS BETAS.

$$
\widehat{\text{price}}
=\widehat\beta_0+\widehat\beta_1(\text{area})
+\widehat\beta_2D_{78721}+\widehat\beta_3D_{78723}.
$$

This model assigns a fixed dollar difference to each additional square foot and to each ZIP comparison. If we take a logarithm of price isntead,  -->


To get some practice with a multiple regression model, we again predict sale price from living area and ZIP code, with 78731 as the reference category:


$$
\widehat{\log(\text{price})}
=\widehat\beta_0+\widehat\beta_1(\text{area})
+\widehat\beta_2D_{78721}+\widehat\beta_3D_{78723}.
$$

Exponentiating the entire fitted equation returns the prediction to dollars and turns the additive components into multipliers:

$$
\begin{aligned}
\widehat{\text{price}}
&=\exp\!\left[
\widehat\beta_0+\widehat\beta_1(\text{area})
+\widehat\beta_2D_{78721}+\widehat\beta_3D_{78723}
\right] \\
&=\exp(\widehat\beta_0)
\exp\!\left[\widehat\beta_1(\text{area})\right]
\exp(\widehat\beta_2D_{78721})
\exp(\widehat\beta_3D_{78723}).
\end{aligned}
$$

A linear regression model makes a prediction by summing up contributions from each term in the model. If we take a logarithm of $Y$ before we fit the linear model, we get a model that makes a prediction by *multiplying* the contributions from each term.

Under normal log-scale errors --- or, more generally, errors symmetric around zero --- this exponentiated fitted value is the conditional median price.

### Checking nonlinearity and equal variance assumptions

Before interpreting the coefficients, we check what the transformation does to the diagnostics. The linear model in @fig-austin-log-diagnostics (left panel) again shows a pronounced increase in residual spread as fitted price rises. After taking logs, the spread is much more consistent, although a hint of nonlinearity remains (it's also visible in the log-price versus area scatterplot above as a leveling-off at higher prices).

```{r}
#| label: fig-austin-log-diagnostics
#| fig-cap: "Residuals versus fitted values for the Austin models. Logging price substantially stabilizes the spread, although the smoother retains a modest bend at high predicted log prices."
#| fig-width: 10
#| fig-height: 4

p_austin_raw_diag <- ggplot(austin_raw_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  scale_x_continuous(labels = dollar_format(scale = 0.001, suffix = "K")) +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "K")) +
  labs(x = "Fitted price", y = "Residual", title = "Price in dollars")

p_austin_log_diag <- ggplot(austin_log_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(
    x = "predicted log price",
    y = "Log-scale residual",
    title = "Log price"
  )

p_austin_raw_diag + p_austin_log_diag
```

<!-- 
None of this is relevant to the pedagogy. 

Adding age, bathrooms, lot size, view, bedrooms, garage spaces, and stories did not flatten the remaining bend during model development. It moved [adjusted $R^2$](regression_02_prediction.qmd#sec-r-squared) only from `r number(austin_base_adj_r2, accuracy = 0.001)` to `r number(austin_rich_adj_r2, accuracy = 0.001)`. We therefore keep the smaller model, since the extra variables do little to address the diagnostic pattern they were meant to repair.

That comparison used the same data to suggest and assess the richer specification. Its small improvement is descriptive, not an independent test of future performance. Adding predictors and transforming the response solve different problems. Extra predictors account for systematic comparisons the model omitted, while logging price changes the form of the outcome and its errors. -->

The log transformation addresses the equal variance violation by changing what "equal variance" means. In the original model, equal variance asks for a similar distribution of errors *in dollars* for any combination of area and ZIP. In the `log(latest_price)` model, it asks for roughly constant **multiplicative** errors across fitted prices. For example, a \$20K error is a 10% miss on a \$200K house, the same percentage miss as \$40K on a \$400K house. In this model, we *expect* larger errors (in dollars) when we predict higher prices.

### Interpreting the coefficients

The estimated coefficients are below:

$$
\widehat{\log(\text{price})}
=`r format_signed(austin_log_coef[["(Intercept)"]], number, accuracy = 0.01, show_plus = FALSE)`
`r format_signed(austin_log_coef[["living_area_sq_ft"]], number, accuracy = 0.00001)`(\text{area})
`r format_signed(austin_log_coef[["zipcode78721"]], number, accuracy = 0.01)`D_{78721}
`r format_signed(austin_log_coef[["zipcode78723"]], number, accuracy = 0.01)`D_{78723}.
$$

@tbl-austin-log-model gives the complete regression table.

```{r}
#| label: tbl-austin-log-model
#| tbl-cap: "Predicting log sale price from living area and ZIP code, with 78731 as the reference category."

coef_table(
  fit_austin_log,
  fmt = 5,
  coef_map = c(
    "(Intercept)" = "Intercept (78731, area = 0)",
    "living_area_sq_ft" = "Living area (sq ft)",
    "zipcode78721" = "78721 vs. 78731",
    "zipcode78723" = "78723 vs. 78731"
  )
)
```

The RSE of `r number(austin_log_rse, accuracy = 0.01)` is measured in log-price units. A positive residual of one RSE corresponds to a price multiplier of `r number(austin_resid_ratio, accuracy = 0.01)`. The observed price is about `r number(100 * (austin_resid_ratio - 1), accuracy = 1)`% above its fitted value. A negative residual of the same size corresponds to a multiplier of `r number(exp(-austin_log_rse), accuracy = 0.01)`. The observed price is then about `r number(100 * (1 - exp(-austin_log_rse)), accuracy = 1)`% below its fitted value. The percentages differ because equal positive and negative changes on the log scale become reciprocal multipliers in dollars.

The table's $R^2$ describes variation in log price. It is not directly comparable with an $R^2$ from a model of price in dollars because the two models measure squared errors on different response scales.


To see how we can interpret these coefficients, consider two homes in ZIP code 78731 with 2,000 and 2,100 square feet. Their predicted log prices are

$$
\begin{aligned}
\widehat{\log(\text{price})}_{2000}
&=\widehat\beta_0+2000\widehat\beta_1
=`r number(austin_area_examples$fitted_log_price[1], accuracy = 0.01)`, \\
\widehat{\log(\text{price})}_{2100}
&=\widehat\beta_0+2100\widehat\beta_1
=`r number(austin_area_examples$fitted_log_price[2], accuracy = 0.01)`.
\end{aligned}
$$

Subtracting the first prediction from the second isolates the change associated with 100 additional square feet:

$$
\widehat{\log(\text{price})}_{2100}
-\widehat{\log(\text{price})}_{2000}
=100\widehat\beta_1
=`r number(area_log_difference, accuracy = 0.0001)`.
$$

Exponentiating that difference gives the *ratio* of the fitted dollar prices:

$$
\frac{\widehat{\text{price}}_{2100}}
{\widehat{\text{price}}_{2000}}
=\exp(`r number(area_log_difference, accuracy = 0.0001)`)
=`r number(area_fitted_ratio, accuracy = 0.001)`.
$$

The 2,100-square-foot home therefore has a fitted price about `r number(100 * (area_fitted_ratio - 1), accuracy = 0.01)`% higher than the 2,000-square-foot home in the same ZIP code. There was nothing special about 2,000/2,100 square feet or choosing 78731; we'd see the same change in prediction for any houses in the same ZIP. You can get to the same interpretation more directly by using the exponentiated regression equation to compute the ratio of fitted prices; anything unchanging between the two predictions cancels out.
<!-- todo, redo w/ dummies retained.
$$
\frac{\widehat{\text{price}}_{2100}}
{\widehat{\text{price}}_{2000}}
=\frac{\exp(\widehat\beta_0+2100\widehat\beta_1)}
{\exp(\widehat\beta_0+2000\widehat\beta_1)}
=\exp(100\widehat\beta_1).
$$
--->

Now compare two 2,000-square-foot homes, one in 78731 and one in 78723. Their predicted log prices are

$$
\begin{aligned}
\widehat{\log(\text{price})}_{78731}
&=\widehat\beta_0+2000\widehat\beta_1
=`r number(austin_zip_examples$fitted_log_price[1], accuracy = 0.01)`, \\
\widehat{\log(\text{price})}_{78723}
&=\widehat\beta_0+2000\widehat\beta_1+\widehat\beta_3
=`r number(austin_zip_examples$fitted_log_price[2], accuracy = 0.01)`.
\end{aligned}
$$

The fitted log-price difference and corresponding dollar-price ratio are

$$
\begin{aligned}
\widehat{\log(\text{price})}_{78723}
-\widehat{\log(\text{price})}_{78731}
&=\widehat\beta_3
=`r format_signed(zip_log_difference, number, accuracy = 0.0001, show_plus = FALSE)`, \\
\frac{\widehat{\text{price}}_{78723}}
{\widehat{\text{price}}_{78731}}
&=\exp(\widehat\beta_3)
=`r number(zip_fitted_ratio, accuracy = 0.001)`.
\end{aligned}
$$

Comparing two houses of the same size, one in 78723 and the other in 78731, the fitted price in 78723 is therefore about `r number(100 * (1 - zip_fitted_ratio), accuracy = 0.1)`% lower than in 78731.

The four coefficients now have distinct interpretations:

- The intercept, $\widehat\beta_0=`r format_signed(austin_log_coef[["(Intercept)"]], number, accuracy = 0.01, show_plus = FALSE)`$, is the predicted log price of a zero-square-foot home in 78731. It's not useful on its own here.
- The area coefficient, $\widehat\beta_1=`r format_signed(austin_log_coef[["living_area_sq_ft"]], number, accuracy = 0.00001, show_plus = FALSE)`$, is the fitted log-price difference per additional square foot among homes in the same ZIP code. A 100-square-foot increase corresponds to a `r number(100 * (area_fitted_ratio - 1), accuracy = 0.01)`% higher fitted price. The 100-square-foot increment is simply to produce a more digestible percentage. It's easy to verify by a similar calculation that a $d$-square-foot increase in area would correspond to a multiplier of $\exp(d\widehat\beta_1)$ on the fitted price.
- The 78721 coefficient, $\widehat\beta_2=`r format_signed(austin_log_coef[["zipcode78721"]], number, accuracy = 0.01, show_plus = FALSE)`$, compares equally sized homes in 78721 with homes in the reference ZIP code, 78731. Exponentiating it gives a fitted price about `r number(abs(zip_78721_pct), accuracy = 0.1)`% lower in 78721.
- The 78723 coefficient, $\widehat\beta_3=`r format_signed(austin_log_coef[["zipcode78723"]], number, accuracy = 0.01, show_plus = FALSE)`$, compares equally sized homes in 78723 with homes in 78731. Exponentiating it gives a fitted price about `r number(abs(zip_78723_pct), accuracy = 0.1)`% lower in 78723.

<!-- 
TOO MANY WORDS NO ONE WILL FOLLOW THIS

The exact conversion follows from comparing two fitted values whose log predictions differ by $b$. Exponentiating their difference gives a fitted-price ratio of $\exp(b)$. Subtracting one turns that ratio into a decimal change, and multiplying by 100 expresses it as a percentage. -->

<!-- 
For living area, a one-square-foot comparison is too small to be useful. Holding ZIP code constant, 100 additional square feet corresponds to `r number(area_100_pct, accuracy = 0.01)`% higher predicted sale price. We calculate that as $100[\exp(100\widehat\alpha_1)-1]\%$ because the coefficient applies once for every added square foot.

The ZIP coefficients compare equally sized homes. Holding area constant, predicted sale price in 78721 is `r number(abs(zip_78721_pct), accuracy = 0.1)`% lower than in 78731; in 78723 it is `r number(abs(zip_78723_pct), accuracy = 0.1)`% lower. The first coefficient is `r format_signed(austin_log_coef[["zipcode78721"]], number, accuracy = 0.001, show_plus = FALSE)`, but exponentiation produces a `r number(abs(zip_78721_pct), accuracy = 0.1)`% decrease. Positive and negative log changes are not symmetric after we return to dollars.

The reference direction matters as well. Saying that 78721 is `r number(abs(zip_78721_pct), accuracy = 0.1)`% lower than 78731 does not mean that 78731 is `r number(abs(zip_78721_pct), accuracy = 0.1)`% higher than 78721. Reversing the comparison uses the reciprocal ratio, $\exp(-\widehat\alpha_2)$, because each percentage uses a different starting value.

The model is additive in log price and multiplicative in dollar price. At any fixed area, moving from 78731 to 78721 multiplies the fitted price by $\exp(`r format_signed(austin_log_coef[["zipcode78721"]], number, accuracy = 0.0001, show_plus = FALSE)`)$. The dollar gap therefore grows with the predicted price rather than remaining fixed. -->


<!--
Technical note for a later treatment: least squares estimates the conditional
mean of log price. Exponentiating that fitted value gives the conditional
geometric mean of price. It also gives the conditional median when the
log-scale errors have median zero, as they do under a symmetric error model.
It does not generally give the conditional arithmetic mean because
exponentiation is nonlinear. Predicting mean dollar price requires a
retransformation correction, such as a model-based correction or a smearing
estimate.
-->

<!--
TODO: ADD BACK WHEN I HAVE TIME TO EDIT THIS MESS

Residuals also take a multiplicative interpretation. For a log-scale residual

$$
e=\log(y)-\widehat{\log(y)},
\qquad
\exp(e)=\frac{y}{\widehat y}.
$$

The exponential of a residual is the ratio of actual price to its dollar-scale fitted value. The Austin model's residual standard deviation is `r number(austin_log_rse, accuracy = 0.01)`, so one residual standard deviation corresponds to a ratio of about `r number(austin_resid_ratio, accuracy = 0.01)`. An error this size is roughly `r number(100 * (austin_resid_ratio - 1), accuracy = 1)`% upward on the dollar scale, not a fixed dollar miss.

A negative residual uses the reciprocal direction. A residual of `r format_signed(-austin_log_rse, number, accuracy = 0.01, show_plus = FALSE)` corresponds to a ratio near `r number(exp(-austin_log_rse), accuracy = 0.01)`, meaning the observed price is about `r number(100 * (1 - exp(-austin_log_rse)), accuracy = 1)`% below its fitted value. The upward and downward percentage readings differ even though the two residuals are equally far from zero on the log scale.

### Four model forms

Logging the response, the predictor, or both produces four related coefficient interpretations. For a slope $b_1$ and a one-unit increase in $X$, the comparison is:

| Model form | Interpretation |
|:---|:---|
| $Y=b_0+b_1X$ | Predicted $Y$ changes by $b_1$ units. |
| $\log(Y)=b_0+b_1X$ | Predicted $Y$ changes by exactly $100[\exp(b_1)-1]\%$, or approximately $100b_1\%$ when $b_1$ is small. |
| $Y=b_0+b_1\log(X)$ | A 1% increase in $X$ corresponds approximately to a $b_1/100$-unit change in predicted $Y$. |
| $\log(Y)=b_0+b_1\log(X)$ | A 1% increase in $X$ corresponds approximately to a $b_1\%$ change in predicted $Y$; $b_1$ is an elasticity. |

: Slope interpretations for linear and logged regression models. {#tbl-log-model-forms}

The last two rows describe small or local changes. For a finite change from $x_1$ to $x_2$, the predictor changes by $\log(x_2/x_1)$, and the response comparison follows from that exact log ratio.

The approximation behaves best when the percentage change is close to zero. If $X$ rises by 1%, then $\log(1.01)=0.00995$, which is close to 0.01. If $X$ rises by 50%, then $\log(1.50)=0.405$ rather than 0.50. An exact finite-change calculation carries useful information here.

--->



<!--PRIORITY ADD BACK NEXT DRAFT!

## Price elasticity {#sec-price-elasticity}

The Kroger data record `r comma(nrow(kroger))` weekly observations of packaged sliced-American-cheese sales across 11 U.S. cities [@kroger-cheese]. Each row gives sales volume in packages, `vol`, and retail price in dollars per package, `price`. The data also record city and whether the product had an in-store display.

A linear model that predicts `log(vol)` from `log(price)` fits a *demand curve* with a familiar form (if you've taken some economics):




allows the predicted change in sales volume associated with a price change to vary with the starting price and volume. Its slope is **price elasticity of demand**: the approximate percentage change in predicted quantity associated with a 1% increase in price.

Elasticity is unit-free because both parts of the comparison are percentages. Measuring price in dollars or cents gives the same slope after the intercept adjusts. The same holds for packages or thousands of packages.

Prices were not randomized, and the fitted model does not account for city or display status. We therefore use its elasticity to describe the fitted price-volume relationship rather than a causal effect of changing price.

@fig-kroger-demand shows both fits on the original axes in the left panel and the straight log-log relationship on logarithmic axes in the right panel.

```{r}
#| label: fig-kroger-demand
#| fig-cap: "Weekly Kroger cheese sales against price. Left: raw linear and log-log fits on the original axes. Right: the log-log relationship is a line on logarithmic axes."
#| fig-width: 10
#| fig-height: 4

p_kroger_original <- ggplot(kroger, aes(price, vol)) +
  geom_point(color = "gray50", alpha = 0.35, size = 1) +
  geom_line(data = kroger_curve, aes(y = raw_fit, color = "Raw linear"), linewidth = 1) +
  geom_line(data = kroger_curve, aes(y = log_log_fit, color = "Log-log"), linewidth = 1.1) +
  scale_color_manual(values = model_colors) +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = comma) +
  labs(x = "Price per package", y = "Packages sold per week", color = "Model")

p_kroger_log <- ggplot(kroger, aes(price, vol)) +
  geom_point(color = "gray50", alpha = 0.35, size = 1) +
  geom_line(
    data = kroger_curve,
    aes(y = log_log_fit),
    color = unname(model_colors["Log-log"]),
    linewidth = 1.1
  ) +
  scale_x_log10(labels = dollar) +
  scale_y_log10(labels = comma) +
  labs(
    x = "Price per package (log scale)",
    y = "Packages sold per week (log scale)"
  )

p_kroger_original + p_kroger_log + plot_layout(guides = "collect")
```

The fitted log-log equation is

$$
\widehat{\log(\text{vol})}
=`r format_signed(kroger_coef[["(Intercept)"]], number, accuracy = 0.0001, show_plus = FALSE)`
`r format_signed(kroger_coef[["log(price)"]], number, accuracy = 0.0001)`\log(\text{price}).
$$

The equation makes log volume linear in log price. The fitted slope is the elasticity defined above.

Exponentiating both sides gives the fitted relationship on the original scale:

$$
\widehat{\text{vol}}
=\exp(\widehat\beta_0)\text{price}^{\widehat\beta_1}.
$$

The intercept sets the curve's overall level, while the exponent determines how predicted volume changes with price. A negative exponent produces the downward curve in @fig-kroger-demand.

Here the fitted elasticity is `r format_signed(kroger_coef[["log(price)"]], number, accuracy = 0.01, show_plus = FALSE)`. In this model, a 1% higher price corresponds to about a `r number(abs(kroger_coef[["log(price)"]]), accuracy = 0.01)`% lower predicted sales volume.

The model gives exact fitted comparisons for larger changes. If price moves from $p_1$ to $p_2$, then

$$
\frac{\widehat{\text{vol}}_2}{\widehat{\text{vol}}_1}
=\left(\frac{p_2}{p_1}\right)^{\widehat\beta_1}.
$$

The fitted volume ratio depends on the price ratio raised to the elasticity. Raising price from \$2.00 to \$2.50 multiplies price by `r number(kroger_price_ratio, accuracy = 0.01)` and gives a fitted volume ratio of `r number(kroger_volume_ratio, accuracy = 0.001)`, or a `r number(100 * (1 - kroger_volume_ratio), accuracy = 0.1)`% decrease. The 1%-change interpretation would be a poor shortcut for a 25% price increase.

The residual plots in @fig-kroger-diagnostics compare the two scales. Logging price and volume changes the residual structure, but the cloud is still messy. Cities, displays, and other omitted factors leave substantial variation.

```{r}
#| label: fig-kroger-diagnostics
#| fig-cap: "Residuals versus fitted values for weekly Kroger cheese sales. The log-log model changes the residual structure but leaves substantial unexplained variation."
#| fig-width: 10
#| fig-height: 4

p_kroger_raw_diag <- ggplot(kroger_raw_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(x = "Fitted weekly volume", y = "Residual", title = "Raw linear model")

p_kroger_log_diag <- ggplot(kroger_log_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(
    x = "Fitted log volume",
    y = "Log-scale residual",
    title = "Log-log model"
  )

p_kroger_raw_diag + p_kroger_log_diag
```

The log-log model has $R^2=`r number(kroger_r2, accuracy = 0.001)`$. Price alone therefore leaves most week-to-week and city-to-city variation unexplained, even though the fitted center of log volume has a consistent downward association with log price. Including city and display status would answer a different adjusted-comparison question.


--->


## Polynomial regression {#sec-polynomial-regression}

**Polynomial regression** allows for nonlinearity by adding additional powers of $X$ to the model. A *degree*-$k$ polynomial regression model has the form

$$
\widehat Y
=\widehat\beta_0+\widehat\beta_1X+\widehat\beta_2X^2+\cdots+\widehat\beta_kX^k.
$$

We treat $X,X^2,\ldots,X^k$ just like any other predictors when we fit the model. A degree-2 polynomial is a quadratic function that can capture $U$-shapes (or nonlinearly increasing or decreasing relationships, depending on where the $X$ values lie). A degree-3 polynomial is a cubic function; it can change direction twice and take more complex forms than a degree-2 polynomial. A degree-$k$ model also includes every lower power, preserving the model's hierarchy.

When we're using polynomials in regression, we usually don't care about the individual coefficients. Instead, we focus on the overall shape of the fitted curve and how well it captures the relationship between $X$ and $Y$.

### Electricity load and temperature

We revisit our running nonlinear example, predicting daily electricity load in North Central Texas from the average daily temperature. Adding a quadratic term captures most of the $U$-shaped relationship that a straight line misses.

```{r}
#| label: fig-ercot-linear-quadratic
#| fig-cap: "Daily mean electricity load against temperature in DFW during 2018. The quadratic fit captures the dominant U-shaped pattern that the straight line misses."
#| fig-width: 10
#| fig-height: 4

p_ercot_degree1 <- ggplot(ercot_2018, aes(temp, load)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_line(
    data = filter(ercot_curves, degree == "1"),
    aes(x, y),
    color = "steelblue",
    linewidth = 1
  ) +
  labs(x = "Daily mean temperature (°F)", y = "Daily mean load (MW)", title = "Degree 1")

p_ercot_degree2 <- ggplot(ercot_2018, aes(temp, load)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_line(
    data = filter(ercot_curves, degree == "2"),
    aes(x, y),
    color = "steelblue",
    linewidth = 1
  ) +
  labs(x = "Daily mean temperature (°F)", y = "Daily mean load (MW)", title = "Degree 2")

p_ercot_degree1 + p_ercot_degree2
```

The fitted curve overlaid on the data suggests that we might be done --- we wanted a $U$ and we got it --- but the residual-versus-fitted plot below shows we still have a violation of the linearity assumption:

```{r}
#| label: fig-ercot-polynomial-diagnostics
#| fig-cap: "Residuals versus fitted values for degree-1 (left) and degree-2 (right) ERCOT models. The quadratic removes the dominant curved pattern."
#| fig-width: 10
#| fig-height: 4

p_ercot_d1_diag <- ggplot(ercot_degree1_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(x = "Fitted load (MW)", y = "Residual (MW)")

p_ercot_d2_diag <- ggplot(ercot_degree2_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(x = "Fitted load (MW)", y = "Residual (MW)")

p_ercot_d1_diag + p_ercot_d2_diag
```

The quadratic captures most of the relationship that the line missed, but some structure remains in the residuals. In particular, for very hot or very cold days (when the fitted values are largest) we are overpredicting demand. You can actually see that in the plot of the model fit over the original data if you look closely. But you can't miss it in the residual versus fitted plot.

@fig-ercot-degree3-check shows the cubic fit and its residual plot.

```{r}
#| label: fig-ercot-degree3-check
#| fig-cap: "The degree-3 ERCOT fit (left) and its residual-versus-fitted plot (right). The cubic makes a smaller adjustment after the quadratic captures the dominant U-shape."
#| fig-width: 10
#| fig-height: 4

p_ercot_degree3 <- ggplot(ercot_2018, aes(temp, load)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_line(
    data = filter(ercot_curves, degree == "3"),
    aes(x, y),
    color = "steelblue",
    linewidth = 1
  ) +
  labs(x = "Daily mean temperature (°F)", y = "Daily mean load (MW)")

p_ercot_d3_diag <- ggplot(ercot_degree3_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(x = "Fitted load (MW)", y = "Residual (MW)")

p_ercot_degree3 + p_ercot_d3_diag
```

The cubic changes the fitted curve only modestly, and the residual smoother still curves downward. Adding a couple more polynomial terms gets us closer to a good fit, as shown in @fig-ercot-degree5-check.

```{r}
#| label: fig-ercot-degree5-check
#| fig-cap: "The degree-5 ERCOT fit (left) and its residual-versus-fitted plot (right). Most remaining departures are small relative to the original nonlinear pattern."
#| fig-width: 10
#| fig-height: 4

p_ercot_degree5 <- ggplot(ercot_2018, aes(temp, load)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_line(
    data = filter(ercot_curves, degree == "5"),
    aes(x, y),
    color = "steelblue",
    linewidth = 1
  ) +
  labs(x = "Daily mean temperature (°F)", y = "Daily mean load (MW)")

p_ercot_d5_diag <- ggplot(ercot_degree5_diag, aes(fitted, residual)) +
  geom_point(color = "gray50", alpha = 0.4, size = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(method = "loess", se = FALSE, color = "darkorange", linewidth = 1) +
  labs(x = "Fitted load (MW)", y = "Residual (MW)")

p_ercot_degree5 + p_ercot_d5_diag
```

<!-- 

```{r}
#| label: tbl-ercot-poly-fit
#| tbl-cap: "Fit measures for polynomial models of 2018 DFW daily mean load on temperature."

knitr::kable(ercot_table, align = c("r", "r", "r", "r"))
```
 -->

### Overfitting and extrapolation

@fig-ercot-poly-curves places the degree-1, degree-2, degree-3, degree-5, and degree-9 polynomials on the same figure. The quadratic captures the main U-shape, with smaller refinements from the other terms. The ninth-degree curve starts producing extra wiggles and behaves erratically at the extremes. This is characteristic of *overfitting*: a model that is so flexible that it can start to chase noise instead of signal.

<!-- 
From degree 1 to degree 2, $R^2$ rises from `r number(ercot_fit_stats$r_squared[ercot_fit_stats$degree == 1], accuracy = 0.001)` to `r number(ercot_fit_stats$r_squared[ercot_fit_stats$degree == 2], accuracy = 0.001)`. Residual SD falls from `r comma(ercot_fit_stats$residual_sd[ercot_fit_stats$degree == 1], accuracy = 1)` MW to `r comma(ercot_fit_stats$residual_sd[ercot_fit_stats$degree == 2], accuracy = 1)` MW. Moving to degree 3 lowers residual SD by only about `r comma(ercot_fit_stats$residual_sd[ercot_fit_stats$degree == 2] - ercot_fit_stats$residual_sd[ercot_fit_stats$degree == 3], accuracy = 1)` MW. -->

```{r}
#| label: fig-ercot-poly-curves
#| fig-cap: "Degree-1, degree-2, degree-3, degree-5, and degree-9 polynomial fits for 2018 DFW daily load. The quadratic captures the dominant U-shape; higher degrees add smaller bends."
#| fig-width: 8.5
#| fig-height: 5

ggplot(ercot_2018, aes(temp, load)) +
  geom_point(color = "gray50", alpha = 0.35, size = 0.9) +
  geom_line(
    data = ercot_curves,
    aes(x, y, color = degree),
    linewidth = 1
  ) +
  scale_color_viridis_d("Degree", option = "D", end = 0.85) +
  labs(x = "Daily mean temperature (°F)", y = "Daily mean load (MW)")
```

High-degree polynomials are especially risky for *extrapolation*. At the low-temperature end of @fig-ercot-poly-curves, the degree-9 curve already bends back down. @fig-ercot-poly-extrapolation extends the temperature range by only eight degrees in each direction and marks the observed boundaries.

```{r}
#| label: fig-ercot-poly-extrapolation
#| fig-cap: "ERCOT polynomial fits extended eight degrees beyond the observed temperature range. Dashed lines mark the data boundaries; the fitted curves separate quickly outside them."
#| fig-width: 8.5
#| fig-height: 5

ggplot(ercot_2018, aes(temp, load)) +
  geom_point(color = "gray60", alpha = 0.25, size = 0.8) +
  geom_line(
    data = ercot_extrap_curves,
    aes(temp, load, color = degree),
    linewidth = 1
  ) +
  geom_vline(xintercept = ercot_temp_range, linetype = "dashed", color = "grey40") +
  scale_color_viridis_d("Degree", option = "D", end = 0.85) +
  labs(x = "Daily mean temperature (°F)", y = "Predicted daily mean load (MW)")
```

The curved models make broadly similar predictions through most of the observed range but diverge rapidly outside it. The extra flexibility that helps a high-degree model follow small in-sample bends also leaves its extrapolations controlled by the highest powers of temperature.

The "Goldilocks" degree is the one that predicts well without chasing noise. Finding it requires checking predictions on new data, which we'll do in the next chapter.

<!---

### Overfitting and extrapolation

The Austin polynomial example uses the full housing file, restricted to ZIP 78744 [@housing-kaggle]. After removing the top 1% of prices within that ZIP code, the sample contains `r comma(nrow(austin_78744))` homes with sale price and age at sale.

The fitted curves in @fig-austin-poly-curves use degrees 1, 2, 3, and 9. The ninth-degree model achieves the highest in-sample $R^2$, but it also develops bends in parts of the age range with relatively few observations.

```{r}
#| label: fig-austin-poly-curves
#| fig-cap: "Sale price versus house age in Austin ZIP 78744, with polynomial fits of degrees 1, 2, 3, and 9. The ninth-degree curve bends sharply near sparse portions of the observed range."
#| fig-width: 8.5
#| fig-height: 5

ggplot(austin_78744, aes(age, price)) +
  geom_point(color = "gray50", alpha = 0.35, size = 0.9) +
  geom_line(
    data = austin_poly_curves,
    aes(x, y, color = degree),
    linewidth = 1
  ) +
  scale_color_viridis_d("Degree", option = "D", end = 0.85) +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "K")) +
  labs(x = "House age at sale (years)", y = "Sale price") +
  coord_cartesian(ylim = range(austin_78744$price))
```

@tbl-austin-poly-fit records the corresponding measures of in-sample fit.

```{r}
#| label: tbl-austin-poly-fit
#| tbl-cap: "Fit measures for polynomial models of sale price on age in Austin ZIP 78744."

knitr::kable(austin_poly_table, align = c("r", "r", "r", "r"))
```

**Overfitting** occurs when a model follows sample-specific variation that need not repeat in new observations. Increasing polynomial degree cannot lower in-sample $R^2$, because the larger model can always reproduce the smaller model by setting its new coefficients to zero. Better in-sample fit therefore does not guarantee better prediction.

@tbl-austin-poly-fit illustrates the mechanical increase. $R^2$ rises from `r number(austin_poly_stats$r_squared[austin_poly_stats$degree == 1], accuracy = 0.001)` for the line to `r number(austin_poly_stats$r_squared[austin_poly_stats$degree == 9], accuracy = 0.001)` for degree 9. Adjusted $R^2$ also rises here. The fitted curve reveals what those summaries omit: extra terms partly improve fit by chasing local rises and dips near sparse ages.

We would need new observations or a validation scheme to measure whether those bends repeat. Without that evidence, the degree-9 model's extra in-sample fit is a possible gain rather than proof of improved prediction. The low-degree curves impose more structure, which can miss a real bend but also makes them less responsive to sample noise.

Those are competing costs. Choosing too little flexibility leaves systematic curvature in the residuals, while choosing too much flexibility can make fitted values unstable and future predictions worse. A low-degree starting point, a residual check, and an honest comparison of the incremental fit keep the choice tied to the relationship at hand.

Polynomial extrapolation creates a sharper danger. @fig-austin-poly-extrapolation extends the degree-2 and degree-9 fits only eight years beyond the largest observed age. The vertical line marks the data boundary; beyond it, the degree-9 curve quickly leaves the range of plausible prices.

```{r}
#| label: fig-austin-poly-extrapolation
#| fig-cap: "Degree-2 and degree-9 Austin fits extended eight years beyond the observed age range. The dashed line marks the oldest observed home; the high-degree curve quickly leaves the plotted price range."
#| fig-width: 8.5
#| fig-height: 5

ggplot(austin_78744, aes(age, price)) +
  geom_point(color = "gray60", alpha = 0.25, size = 0.8) +
  geom_line(
    data = austin_extrap_grid,
    aes(age, price, color = degree),
    linewidth = 1.1
  ) +
  geom_vline(xintercept = austin_max_age, linetype = "dashed", color = "grey40") +
  annotate(
    "text",
    x = austin_max_age,
    y = 575000,
    label = "Observed range ends",
    hjust = 1,
    size = 3.2
  ) +
  scale_color_viridis_d(option = "D", end = 0.85) +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "K")) +
  coord_cartesian(xlim = c(0, 72), ylim = c(0, 650000)) +
  labs(x = "House age at sale (years)", y = "Predicted sale price", color = NULL)
```

By age 68, the degree-9 model predicts `r dollar(austin_degree9_at_68, accuracy = 100)`, an impossible negative price. The line drops outside the plotted window. The quadratic remains visually restrained, but its extrapolation still depends on an untested continuation of the fitted curve.

Every polynomial eventually follows its highest power, so even a smooth low-degree fit can behave poorly outside the observed range. High degrees can make that instability arrive faster. In this introductory setting, start with a low degree when it materially repairs a missed curve. Residuals then show whether the repair worked. High-degree wiggles require stronger justification, and extrapolation remains risky at every degree.

The extrapolation problem differs from ordinary uncertainty inside the data range. Within the observed ages, nearby homes constrain the curve. Beyond age `r number(austin_max_age, accuracy = 1)`, no observations oppose the rapidly growing ninth-power term, so its direction comes from algebra rather than nearby evidence.

Log transformations and polynomial terms address different problems. Logging changes the scale of the response or predictor and turns coefficients into multiplicative comparisons. Polynomial terms bend the fitted relationship while leaving the response in its original units.

---> 

## Appendix: Logarithms and exponentials {.unnumbered}

The natural logarithm $\log(x)$ is defined only for $x>0$. It is the inverse of the exponential function $\exp(x)=e^x$, so

$$
\log(1)=0,\qquad \exp(0)=1,\qquad
\log(\exp a)=a,\qquad \exp(\log x)=x.
$$

The log and exponential functions undo one another. The last identity requires $x>0$.

Logs turn products into sums, ratios into differences, and powers into multiplication:

$$
\log(ab)=\log a+\log b,
\qquad
\log(a/b)=\log a-\log b,
\qquad
\log(a^k)=k\log a.
$$

Each rule converts a multiplicative operation on the original scale into an additive operation on the log scale. Exponentials reverse the same move, with $\exp(a+b)=\exp(a)\exp(b)$.

A difference between two logged values is the log of their ratio:

$$
\log(y_2)-\log(y_1)=\log(y_2/y_1).
$$

A log difference $d$ therefore corresponds to the exact ratio $\exp(d)$ and the exact percentage change $100[\exp(d)-1]\%$. For small decimal changes, $\log(1+r)\approx r$ and $\exp(d)-1\approx d$. A 2% increase has $\log(1.02)=0.0198$, so the approximation differs by only 0.0002. A 50% increase has $\log(1.50)=0.4055$, so the same shortcut is much rougher.

The sign carries the direction. Positive log differences produce ratios above one, zero produces a ratio of one, and negative log differences produce ratios between zero and one. Equal positive and negative log differences correspond to reciprocal ratios: $\exp(0.2)=1.22$ and $\exp(-0.2)=0.82$.

Natural logs use base $e\approx2.718$, but the base does not change the substantive pattern once coefficients are interpreted consistently. We use natural logs because their derivatives give the percentage-change approximations above and because statistical software commonly treats `log` as the natural logarithm.

Zero and negative values cannot be logged. Adding an arbitrary constant changes ratios and therefore changes the scientific meaning of the analysis, so it is not a generic fix for nonpositive observations.

## Data used in this section {.unnumbered}

- [Austin House Listings](appendix_data.qmd#sec-data-housing-kaggle)
- [ERCOT Hourly Load and Temperature Panel](appendix_data.qmd#sec-data-ercot-shaffer)
