---
title: "Measuring Location and Spread"
---

```{r}
#| label: setup
#| include: false
#| echo: false
library(tidyverse)
library(scales)
library(patchwork)

# Figures render at their intrinsic size (fig-width x 96px), capped by the
# ~702px content column -- they are never upscaled to fill it. So axis text
# lands at its true point size, and base_size, not fig-width, is what makes it
# legible next to 17px body text. At 13, tick labels are ~14px.
theme_set(theme_minimal(base_size = 13))

austin_full <- read_csv("data/austin_houses/austin_houses.csv",
                        show_col_types = FALSE)

austin_zip <- austin_full |>
  filter(zipcode == 78752)
```

Visualizations show the shape of a distribution, but we often need concise, numerical descriptions of our data. Statistical summaries compress a distribution into a few quantities which we can report, compare, or use in subsequent statistical analyses.

## Measures of center {#sec-measures-center}

The two most common measures of center are the **mean** and the **median**.

The **mean** is the arithmetic average: add up all the values and divide by the number of observations. For our Austin 78752 prices, the mean is `r dollar(round(mean(austin_zip$latest_price)))`. In notation, we write $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$ --- the sum of all values divided by the count.

The **median** (aka the 50th percentile, 0.5 quantile, or second quartile) is the middle value when the data are sorted from smallest to largest. For our prices, the median is `r dollar(round(median(austin_zip$latest_price)))`.

:::: {.TechnicalDetail}
## Medians when values tie

When the number of observations is odd, the median is the middle sorted value. When it is even, we use the average of the two middle values. Ties can mean that fewer than half of the observations fall strictly below or strictly above the median. The general rule is that at least half are at or below the median and at least half are at or above it.
::::

The mean and median are close to each other for symmetric distributions. When a distribution is right-skewed, the mean gets pulled toward the tail by extreme values. That's what we see here: the mean (`r dollar(round(mean(austin_zip$latest_price)))`) is higher than the median (`r dollar(round(median(austin_zip$latest_price)))`) because a few high-priced homes drag the average up. The median, by contrast, is unaffected by how extreme the extremes are --- it only cares about the midpoint of the sorted data. It's a good idea to report both; substantial differences between them indicate the presence of outliers, skewness, or other asymmetries in the distribution.

Consider the bimodal distribution of house ages in the full Austin housing dataset (@fig-house-age-center).

```{r}
#| label: fig-house-age-center
#| fig-cap: "Distribution of house ages in the full Austin housing dataset."
ggplot(austin_full, aes(x = age)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "steelblue", color = "white") +
  geom_density(alpha = 0.3) +
  geom_rug(alpha = 0.1) +
  labs(x = "House age (years)", y = "Density")
```

The distribution has two broad peaks, while the mean age of `r round(mean(austin_full$age), 1)` years falls between them. A single measure of center hides that structure.


## Measures of spread {#sec-measures-spread}

Knowing the center of a distribution isn't enough --- we also need to know how much values vary around that center. We can think about the spread of a distribution as measuring how *un*predictable a variable is. Let's see an example. Below we plot the distribution of sale prices in ZIP codes 78741 and 78749 from the full Austin housing dataset.

```{r}
#| label: fig-zip-comparison
#| fig-cap: "Sale prices by ZIP code: box plots (left) with x's indicating sample means and overlapping density-scaled histograms (right)."
#| fig-width: 9
#| fig-height: 4
zip_data <- austin_full |>
  filter(zipcode %in% c(78749, 78741)) |>
  mutate(zipcode = factor(zipcode))

p_box <- ggplot(zip_data, aes(x = zipcode, y = latest_price)) +
  geom_boxplot(fill = "steelblue", alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 4, size = 3, color = "darkred") +
  scale_y_continuous(labels = dollar_format()) +
  labs(x = "ZIP code", y = "Sale price")

p_hist <- ggplot(zip_data, aes(x = latest_price, fill = zipcode)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 20, color = "white", alpha = 0.6,
                 position = "identity") +
  geom_rug(aes(color = zipcode), alpha = 0.3, show.legend = FALSE) +
  scale_x_continuous(labels = dollar_format()) +
  labs(x = "Sale price", y = "Density", fill = "ZIP code")

p_box + p_hist
```

Now imagine you need to guess the sale price for a new listing in each of those ZIP codes, using only the ZIP code to make your guess. Assuming our historical data is useful for making such a prediction, which price is going to be harder to accurately predict?


You probably said 78741, because the values are more "spread out" than in 78749. Your best prediction for the sale price of a house in either ZIP code will be the mean or median (depending on how you define "best" mathematically). Regardless of which measure of center we choose, there is more natural variability in 78741 than in 78749. We could also say that the sale price of a new house in 78741 is more *unpredictable* than in 78749, absent new information beyond the ZIP code. Assuming this history is representative of future sales, our prediction error will tend to be smaller in 78749.

```{r}
#| label: spread-setup
#| include: false
p49 <- austin_full$latest_price[austin_full$zipcode == 78749]
p41 <- austin_full$latest_price[austin_full$zipcode == 78741]
```

:::: {.Definition #defn-standard-deviation}
## Standard deviation

The **standard deviation** measures the typical distance of an observation from the mean. We square the deviations so that positive and negative ones don't cancel out, then take the square root to get back to the original units.

::: {.details .formula-details #formula-standard-deviation summary="Details and formulas" non-interactive-summary="Formula"}

$$
s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n} (x_i - \bar{x})^2}
$$

The standard deviation is the square root of the average squared distance from the mean. Dividing by $n-1$ rather than $n$ to form that average is a technical adjustment; in any case we care about, $n$ and $n-1$ are effectively the same.
:::
::::

For 78749 the standard deviation is `r dollar(round(sd(p49)))`; for 78741 it is `r dollar(round(sd(p41)))`. That gap matches what we saw: the standard deviation in 78741 is about 1.8 times the standard deviation in 78749.

:::: {.Definition #defn-variance}
## Variance

The **variance** is simply the standard deviation squared.
::::

For 78749 that comes to $s^2 = `r format(round(var(p49) / 1e9, 1), big.mark = ",")`$ billion dollars-squared, and for 78741 about $`r format(round(var(p41) / 1e9, 1), big.mark = ",")`$ billion. The variance is most useful as a measure of spread --- unpredictability --- when we are considering multiple variables at once, because it decomposes into pieces that can be attributed to different sources.

The **interquartile range** (IQR) is the distance between the 75th and 25th percentiles, $\text{IQR} = Q3 - Q1$ --- the width of the middle half of the data. It works out to `r dollar(round(IQR(p49)))` in 78749 and `r dollar(round(IQR(p41)))` in 78741. Like the median, the IQR ignores how extreme the extremes are, which makes it a more stable summary of spread than the standard deviation when a distribution is skewed or carries outliers.

## Data used in this section {.unnumbered}

- [Austin House Listings](appendix_data.qmd#sec-data-housing-kaggle)
